---
title: "npmatch: Building a Stage-1 Match Classifier"
format:
  html:
    df-print: kable
    highlight-style: tango
    toc: true
    toc-location: right
    css: clean.css
    embed-resources: true
execute:
  echo: true
  eval: false
  warning: false
vignette: >
  %\VignetteIndexEntry{npmatch: Building a Stage-1 Match Classifier}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

*How to turn the labeled benchmark into a Stage-1 classifier that decides
"is this candidate the right EIN?" — and, more usefully, how extending the
training data with extra BMF/SAM fields and a few feature-engineering steps
improves on the hand-tuned scoring weights. The lessons matter more than the
model: **gate degeneracy, model gradation, and let a small tree tell you which
is which.** All results below are from `TRAINING-PAIRS-FULL` (304,274 scored
candidate pairs; 1,168 true matches).*

## 1. The task, and the trap

The classifier scores candidate pairs, but the real job is per **query**: for
each source org (`uei`), pick the one correct `ein` — or none. Two facts shape
everything:

- **The pairs are already blocked candidates.** So evaluate *query-level*
  (did we pick the right EIN?), not per-pair.
- **The label is 0.38% positive** (1,168 / 304,274). At that base rate a model
  that always predicts "no match" is 99.6% accurate. **Accuracy is a trap** —
  lead with precision/recall and the query-level selection rate, and either
  down-sample negatives or class-weight when fitting.

```{r}
library(data.table); library(rpart)
pairs <- as.data.table(read.csv(
  np_fetch(np_source_urls()$url[np_source_urls()$key == "training_pairs_full"])))
# match_yn = 1 if this ein is the ground-truth match for the uei
```

## 2. Feature engineering

The raw similarities are the backbone; a handful of engineered features attack
the specific failure modes of pure string similarity.

```{r}
pairs[, name_exact    := as.integer(name_sim >= 0.999)]        # the jump at 1.0 is qualitative
pairs[, log_name_freq := log1p(name_freq)]                     # distinctiveness (rare name = trustworthy)
pairs[, subset_flag   := as.integer(name_match_type == "token_overlap")]  # one name inside the other
pairs[, group_member  := as.integer(BMF_group_exemption_is_member %in% c("TRUE","1"))]
# officer / contact-name agreement (continuous, not 0/1)
pairs[, officer_sim := ifelse(nzchar(BMF_in_care_of) & nzchar(SAM_poc_last_name),
        stringdist::stringsim(toupper(BMF_in_care_of), toupper(SAM_poc_last_name), "jw", p = .1), NA)]
# name-token conflicts (a "Fund II" vs "Fund III" is a strong negative)
conflict <- function(a, b) as.integer(nzchar(a) & nzchar(b) & a != b)
pairs[, `:=`(num_conflict = conflict(name_nums_x, name_nums_y),
             ord_conflict = conflict(name_ord_x,  name_ord_y),
             dir_conflict = conflict(name_dir_x,  name_dir_y))]
```

Why these, and not others:

| Feature | Rationale |
|---|---|
| `name_sim` **and** `name_exact` | the relationship is non-linear — an exact match is a different regime, and trees split on it cleanly |
| `log_name_freq` | *distinctiveness*: an exact match on a unique name is trustworthy; on "First Baptist Church" it is nearly worthless. Interact it with `name_sim` |
| `geo_stnum`, `geo_zip5`, `street_sim` | address agreement; `geo_stnum` (same street number) is the strongest single address signal |
| `subset_flag` | catches truncation/expansion that Jaro-Winkler misses ("Lake Stewards" ⊂ "Lake Stewards of Maine") |
| `group_member` | federated chapters (Habitat, Rotary, VFW) share names → inherently ambiguous → the model should down-weight confidence |
| `officer_sim` | a matching director / point-of-contact name disambiguates small orgs with generic names |

We deliberately **exclude** `score` (the current model's own output — including it
just teaches "trust the old scorer"), `pass`, and `is_top_candidate` (the algo's
pick — direct leakage). They are baselines to beat, not inputs.

## 3. Vetoes vs. features — gate degeneracy, model gradation

Before modeling, ask of each categorical region: *what does P(match) look like
here?* If it is a near-deterministic **zero** with a mechanism, it is a **rule**
(filter it out — this de-noises training and cuts the imbalance). If it merely
shifts the probability, it is a **feature**. The test is the **upper confidence
bound** on the region's match rate — not the point estimate — plus a mechanism
and enough support:

```{r}
wilson_u <- function(pos, n, z = 1.96) { p <- pos/n
  (p + z^2/(2*n) + z*sqrt((p*(1-p) + z^2/(4*n))/n)) / (1 + z^2/n) }
pairs[, .(n = .N, matches = sum(match_yn), rate = mean(match_yn),
          upper95 = wilson_u(sum(match_yn), .N)), by = .(SAM_bus_type_foreign_owned)]
```

Run over the candidate cells, this is what the benchmark shows:

| Cell | n | matches | rate | **95% upper** | verdict |
|---|---:|---:|---:|---:|---|
| SAM `foreign_owned` | 3,643 | **0** | 0.00% | **0.11%** | **→ veto** (large, zero, mechanism) |
| SAM `for_profit` | **3** | 0 | 0.00% | **56%** | **✗ not a veto** — n too small to conclude |
| BMF group member | 38,577 | 63 | 0.16% | 0.21% | feature (down-weight) |
| SAM struct `2L` (not tax-exempt) | 17,747 | 31 | 0.18% | 0.25% | feature |
| SAM `nonprofit` | 264,032 | 1,135 | 0.43% | — | base rate |
| name exact (`sim`=1) | 1,465 | 996 | 68% | 70% | dominant signal |

**The lesson in one table.** `foreign_owned` and `for_profit` have the *same*
point estimate (0%), but only `foreign_owned` is a veto: n = 3,643 gives an upper
bound of 0.1%, and an international org cannot have a domestic BMF record (the
mechanism guarantees it generalizes). `for_profit` has n = 3 and an upper bound
of 56% — you cannot conclude anything, and promoting it to a rule would be
overfitting. This is why small pure cells must never become rules on the point
estimate alone, and why the mechanism check also guards against filtering on a
blocking artifact (these are *already* candidate pairs).

Promote only `foreign_owned` (and the existing hard vetoes) to the gate; leave
`group_member`, structure, and business-type as features the model can grade.

## 4. A small, interpretable tree

Fit a shallow, pruned CART on a grouped, down-sampled training split (all
positives + ~15× negatives; queries never split across folds):

```{r}
ueis <- unique(pairs$uei); tr_uei <- sample(ueis, floor(.7 * length(ueis)))
tr <- pairs[uei %in% tr_uei]; te <- pairs[!uei %in% tr_uei]
pos <- tr[match_yn == 1]; neg <- tr[match_yn == 0][sample(.N, 15 * nrow(pos))]
fit <- rpart(match_yn ~ name_sim + name_exact + log_name_freq + geo_stnum + geo_zip5 +
             geo_state + street_sim + city_sim + name_match_type + group_member +
             subset_flag + officer_sim + num_conflict,
             data = rbind(pos, neg), method = "class",
             control = rpart.control(maxdepth = 4, cp = .004, minbucket = 40))
fit <- prune(fit, cp = fit$cptable[which.min(fit$cptable[, "xerror"]), "CP"])
print(fit)
```

The pruned tree is three leaves:

```
name_sim < 0.94 ?
├─ yes → street_sim < 0.88 ?
│        ├─ yes → NO   (0.2% match)   the clean-negative leaf
│        └─ no  → YES  (77% match)    fuzzy name rescued by a matching street
└─ no  → YES  (95% match)
```

A **two-variable story with an interaction**: a strong name is enough on its own;
a weak name is rescued only when the street agrees (the address-anchored
recovery); everything else is a non-match. Variable importance echoes it —
`name_sim` (.29), `name_exact` (.26), `street_sim` (.17), `geo_stnum` (.16),
`geo_zip5` (.12) — roughly five features carry all the signal. This is the
tree-as-search idea in action: it *discovers* where the effect of `name_sim`
varies (its threshold differs by branch), i.e. it gives you varying slopes
without saturating a factor table.

## 5. Does it work? Two lessons

Pick the highest-scoring candidate per query and ask whether it is the true match,
against the hand-tuned hier score as the baseline (queries split by `uei`,
`foreign_owned` gated out):

| Selector | ALL | RANDOM | **HARD** |
|---|---:|---:|---:|
| Hand-tuned hier score | 0.934 | 0.961 | 0.877 |
| Small 3-leaf tree | 0.915 | 0.927 | 0.891 |
| **Logistic regression (core features)** | **0.945** | 0.961 | **0.912** |

**Lesson 1 — trees explain; smooth models select.** The three-leaf tree is a
superb *summary* (two variables, one interaction, a clean veto) and lands close,
but it *ranks* coarsely: a shallow tree assigns each candidate one of a few leaf
probabilities, so many candidates in a query tie and the within-query argmax is
arbitrary. That is why the tree sits below the continuous score. Use the tree for
interpretation and veto discovery; use a smooth model for the actual selection.

**Lesson 2 — a learned model beats the hand-tuned weights, most where it matters.**
A plain logistic regression on the *same* core features tops the hand-tuned score
overall (0.945 vs 0.934) and by **+3.5 points on the hard cases** (0.912 vs
0.877). Learning the weights from the labeled data — rather than setting
`0.6·name + 0.4·geo` by hand — is the improvement, and it concentrates exactly on
the difficult residual.

```{r}
m <- glm(match_yn ~ name_sim + name_exact + geo_stnum + geo_zip5 + geo_state +
         street_sim + city_sim, family = binomial, data = rbind(pos, neg))
sort(round(coef(m), 2))
```

And the coefficients are legible — they *are* the story the tree told, quantified:

```
name_sim  +21.0   geo_stnum  +4.1   name_exact  +3.5   geo_state  +2.4   geo_zip5  +1.1
                                                       (city/street positive; large negative intercept)
```

**The honest twist on feature engineering.** Adding the engineered features
(`log_name_freq` interaction, `group_member`, `subset_flag`, `officer_sim`, the
conflict flags) to the logistic model did **not** move top-1 selection on this
benchmark — the core name+geo signal already carries it. Their value shows up
elsewhere, and it is real: their **coefficients confirm every mechanism** —
`num_conflict` ≈ −13 (a "Fund II" vs "Fund III" all but rules the pair out),
`group_member` ≈ −2 (federated chapters are genuinely weaker, your #10),
`name_sim:log_name_freq` ≈ −2 (an exact match on a *common* name is worth less —
the distinctiveness interaction) — and they sharpen **precision on the ambiguous
tail**, not the headline metric. The lesson is the discipline you asked for:
**don't pile on features reflexively.** Learn the weights (that is the win), let a
tree show you the few variables and where the slopes vary, gate the one
mechanistic dead-zone (`foreign_owned`), and add engineered signals where a
*mechanism* — not just a hoped-for lift — says they belong.

## 6. From classifier to pipeline

The gate-then-model split you just built is exactly the package's architecture:
the veto layer (`np_veto`) removes mechanistically-impossible pairs; the scorer
(`np_score`) grades the rest; `np_select` / `np_tier` turn per-pair scores into a
per-query YES / MAYBE / NO. A learned Stage-1 model drops straight into the scorer
slot, and the MAYBE tier still routes the uncertain middle to the LLM review loop
(`np_as_prompts`). Retrain it as the benchmark grows — the training pairs are a
projection of the evaluation frame, so every improvement is reproducible.

## Appendix — model menu

The main text uses one model on purpose. For accuracy or a second opinion:

| Model | R package | Note |
|---|---|---|
| Pruned CART | `rpart` (+ `rpart.plot`) | the interpretable centerpiece used above |
| C5.0 / cond-inference tree | `C50`, `partykit::ctree` | `ctree`'s significance-tested splits won't manufacture a spurious pure cell — good for veto mining |
| Rule learner (RIPPER) | `RWeka::JRip` | targets "high-precision rule over a region" directly — the veto-mining objective as an algorithm |
| Random forest | `ranger` | fast accuracy benchmark + variable importance |
| Gradient boosting | `xgboost` | max accuracy; use `scale_pos_weight` for the imbalance |
| Logistic regression | base `glm`, `npmatch::np_train` | a parametric baseline; add varying slopes only on the few factor×continuous cells the tree flags |
| Optimal classification tree | `iai` (commercial) / `evtree` (open) | marginal gain over pruned CART here — not required |

Wrap any of these in `tidymodels` (`recipes` for the engineering, `rsample` for
**grouped** CV by `uei`, `yardstick` for PR-AUC) to keep the comparison honest.
