---
title: "The Candidate Evaluation Frame — Fields & Methods"
format:
  html:
    theme: readable
    df-print: kable
    highlight-style: tango
    toc: true
    toc-location: left
    css: clean.css
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{The Candidate Evaluation Frame — Fields & Methods}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(npmatch)
have <- function(p) requireNamespace(p, quietly = TRUE)
```

*A technical companion to the Reviewer's Guide. It defines every field in the
evaluation frame npmatch returns and explains how the derived measures —
similarity scores, token weights, name-distinctiveness, and the geographic
hierarchy — are computed.*

## What the evaluation frame is

The matcher returns **one row per surfaced candidate** (a source organization
with several plausible EINs contributes several rows). Each row is a
self-contained *review packet*: everything a human or an LLM needs to judge
whether that candidate is the right match, arranged in an **information
hierarchy** from the bottom line up:

1. **Match strength & outcome** — the score, the decision, and any veto.
2. **Name-match summary** — which name matched, how, and how distinctive it is.
3. **Name-cleaning progression** — raw → normalized → tokenized, for both sides.
4. **Address & geography** — field similarities and exact-agreement flags.
5. **Imported context** — descriptive fields carried verbatim from each source,
   flagged by a `SAM_` / `BMF_` prefix.

Two conventions run throughout: a **`_uss` / `_bmf` suffix** marks which side a
field describes (source vs. reference), and a **`SAM_` / `BMF_` prefix** marks a
field *imported* from that dataset rather than *derived* by npmatch.

## Field dictionary

### 1 · Match strength & outcome
```{r d1}
data.frame(Field = c("uei / ein","name_similarity","addr_similarity",
  "candidate_type","total_score","num_of_candidates","is_top_candidate",
  "match_decision","match_layer","decision_reason","veto / veto_reason",
  "veto_soft / veto_soft_reason","notes"),
  Meaning = c(
  "the two entity keys — source UEI and reference EIN",
  "0–1 agreement of the names (see Methods)",
  "0–1 agreement of the addresses (best available granularity)",
  "how this candidate entered — exact, token, de-spaced, cross-state",
  "combined 0–1 score = 0.6·name_similarity + 0.4·geo (see Methods)",
  "how many candidates were surfaced for this source org (review group size)",
  "1 on the selected pick (YES/MAYBE); 0 on the others",
  "the tier: YES (auto-accept) / MAYBE (review) / NO (reject)",
  "which cascade pass produced this candidate",
  "short human-readable reason for the tier",
  "hard do-not-match rule fired → forces NO (e.g. for-profit legal form)",
  "soft rule fired → capped at MAYBE (e.g. affiliate suffix, government entity)",
  "free-text annotations"), check.names = FALSE)
```

### 2 · Name-match summary
```{r d2}
data.frame(Field = c("match_name_uss / match_name_bmf",
  "match_version_uss / match_version_bmf","match_type","normalized_match_count"),
  Meaning = c(
  "the actual name string that produced the match, on each side",
  "which name version matched: MAIN, DBA, or DIVISION",
  "exact · name (fuzzy) · dba · token_overlap — how the names were matched",
  "how many reference records share this exact normalized name — a distinctiveness measure (see Methods)"),
  check.names = FALSE)
```

### 3 · Name-cleaning progression (each side, `_uss` and `_bmf`)
```{r d3}
data.frame(Field = c("name_*_raw_main","name_*_raw_dba","name_*_raw_division",
  "name_*_normalized","name_*_org_type","name_*_tokenized"),
  Meaning = c(
  "the original legal name as supplied",
  "the original DBA / alternate name, if any",
  "the original division / second alternate name, if any",
  "the cleaned match key (suffixes, 'The', punctuation removed; abbreviations standardized)",
  "the organization-type words that were stripped (Foundation, Inc, Association ...)",
  "the informative tokens with their IDF weights, rarest first — TOKEN(idf) (see Methods)"),
  check.names = FALSE)
```

### 4 · Address & geography
```{r d4}
data.frame(Field = c("street_similarity / city_similarity / zip_similarity",
  "street_* / city_* / state_* / zip5_*  (per side)","street_*_normalized",
  "geo_stnum / geo_zip9 / geo_zip5 / geo_zip3 / geo_state / geo_pobox"),
  Meaning = c(
  "0–1 field-level agreement for street, city, ZIP",
  "the raw address components carried for both sides",
  "the standardized street body used for comparison",
  "exact-agreement flags (1/0) at each granularity of the ZIP/geo hierarchy (see Methods)"),
  check.names = FALSE)
```

### 5 · Imported context (appended last)
`SAM_`-prefixed fields (entity structure, NAICS, points of contact, ...) and
`BMF_`-prefixed fields (NTEE code, 501(c) subsection, ruling year, assets,
revenue, foundation status, ...) are carried **verbatim** from the source and
reference datasets to give the reviewer background. They are not computed by
npmatch and are documented in their originating dataset's data dictionary; the
prefix is the provenance flag.

## Methods — how the derived measures are computed

### Name similarity (Jaro–Winkler)
Name agreement is a **Jaro–Winkler** similarity in [0, 1]. Jaro counts how many
characters the two strings share **within a sliding window** (so order and small
gaps are tolerated) and penalizes **transpositions** (characters matched but out
of order). **Winkler** then adds a *prefix boost*: strings that agree on their
first few characters score higher, because shared beginnings are strong evidence
for organization names. npmatch uses a prefix factor `p = 0.1` and **floors**
similarities below `jw_threshold` (default 0.85) to 0, so unrelated strings
contribute nothing.

```{r jw}
if (have("stringdist")) {
  ex <- data.frame(
    A = c("MERCY HEALTH","MERCY HEALTH","SAINT MARYS HOSPITAL","STEP FORWARD"),
    B = c("MERCY HEALTH SYSTEM","MEDSTAR HEALTH","SAINT MARYS HOSP","STEPFORWARD"))
  ex$jaro         <- round(stringdist::stringsim(ex$A, ex$B, method = "jw", p = 0),   3)
  ex$jaro_winkler <- round(stringdist::stringsim(ex$A, ex$B, method = "jw", p = 0.1), 3)
  ex
}
```

The `jaro_winkler` column is what npmatch uses. Note how the shared prefix lifts
"MERCY HEALTH" / "MERCY HEALTH SYSTEM" above the unrelated "MEDSTAR HEALTH", and
how a de-spaced or truncated variant still scores high. Where the character-level
score misses a real match — reordered words, acronyms, subset/superset names —
a **token-overlap** measure takes over (that match is capped at MAYBE, since
containment alone can't separate a parent from a subsidiary).

### Tokenized names and IDF weights
A normalized name is split into **tokens** (words). Each token carries an
**inverse document frequency (IDF)** weight, `log(N / d)`, where `N` is the
number of reference records and `d` is how many contain that token. **Rare tokens
score high** (they pin down an organization); common ones score low. The
`name_*_tokenized` field lists tokens **rarest-first** as `TOKEN(idf)`, so a
reviewer can see at a glance which words carry the match.

```{r idf}
ref <- c("COMMUNITY HEALTH CENTER","COMMUNITY ACTION AGENCY","MERCY COMMUNITY CLINIC",
         "STEPFORWARD","MERCY HEALTH","HEALTH FOUNDATION","COMMUNITY FOUNDATION")
idf <- np_token_idf(ref)
data.frame(token = names(idf), idf = round(as.numeric(idf), 2))[order(-idf), ][1:6, ]
```

Here "STEPFORWARD" (appears once) is far more discriminating than "COMMUNITY"
(appears in many records). Blocking uses these weights so a pair is only proposed
when it shares *distinctive* words — one rare token, or several moderately common
ones, not a lone "COMMUNITY".

### Name distinctiveness (`normalized_match_count`)
This counts how many reference records share a candidate's exact normalized name.
A count of **1** means an exact name match is trustworthy on its own (few things
are named that); a large count (e.g. "First Baptist Church") means an exact name
match needs address corroboration. npmatch uses it to decide when an exact name
alone justifies a confident tier.

### The geographic hierarchy and the geo score
ZIP codes form a **nested prefix family**: the 3-digit region contains the
5-digit ZIP, which the +4 refines to `zip9`.

> `zip3` ⊃ `zip5` ⊃ `zip9`  ·  plus street number, PO box, city, state

The `geo_*` flags record exact agreement at each level. The **geo sub-score takes
the strongest confirmed granularity via a `max`, not a sum** — so a matching
ZIP+4 (or street) counts fully, correlated fields aren't double-counted, and a
wrong *city* can't drag down a *ZIP*-confirmed match. Granularity weights,
strongest to weakest:

```{r geo}
gw <- np_default_geo_weights()
data.frame(granularity = names(gw), weight = as.numeric(gw))
```

This is what lets the score **degrade gracefully**: an org known only to the
state level still scores (weakly) on geography, while one confirmed to the
building scores fully — without either being penalized for the detail it lacks.

### The combined score
```
total_score = 0.60 · name_similarity  +  0.40 · geo_score
```
with `geo_score` the max-over-granularity value above. A distinctive exact name
can be promoted (trusted without address); a token-overlap-only match is capped
below YES. The `match_decision` tier then applies the YES (≥ 0.78) and MAYBE
(≥ 0.65) thresholds, plus a runner-up **margin** check so a near-tie between two
candidates routes to review rather than auto-accepting the wrong one.

## Appendix — field order in the frame

Fields are emitted in the information hierarchy above: keys and decision first,
then the name summary, the name-cleaning progression, address and geography, and
finally the imported `SAM_` / `BMF_` context. The `_uss` / `_bmf` suffix always
marks the side; the `SAM_` / `BMF_` prefix always marks an imported (non-derived)
field.
