---
title: "npmatch: A Reviewer's Guide"
format:
  html:
    theme: readable
    df-print: paged
    highlight-style: tango
    toc: true
    toc-location: left
    css: clean.css
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{npmatch: A Reviewer's Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.2, dpi = 110)
library(npmatch)
have <- function(p) requireNamespace(p, quietly = TRUE)
if (have("ggplot2")) library(ggplot2)
```

*A plain-language overview of how the nonprofit crosswalk is built, what the
package decides on its own, and where a human — or an LLM standing in for one —
comes in.*

## 1. What the package does

**npmatch** links organizations from one list (here, federal award and
registration records keyed by a **UEI**) to the IRS **Business Master File**
of nonprofits (keyed by an **EIN**). It does three things:

1. **Normalizes** both sides so they can be compared fairly.
2. **Runs a multi-tier matching strategy** — strict first, then progressively
   looser — scoring how well each candidate agrees on name and address.
3. **Returns a scored, labeled candidate dataset** — for every source
   organization: the candidate EIN(s) it weighed, a 0–1 match score, and a
   **decision label: YES, MAYBE, or NO**.

The output is not just "the answer." It is an auditable table showing *which
candidates were considered and why one was chosen* — the raw material for the
review stage.

## 2. How names and addresses are normalized

Before anything can be compared, "The Smith Family Foundation, Inc." and
"Smith Family Fdn" have to be made to look alike. npmatch cleans each name to a
**match key**, then splits it into **tokens** (the words the matcher actually
compares). Legal suffixes (Inc, Foundation), a leading "The", and punctuation
are dropped; abbreviations are standardized (St → Saint, & → and, Ctr → Center).

```{r norm-names}
nm <- c("The Smith Family Foundation, Inc.", "St. Mary's Hosp. & Health Ctr",
        "Boys & Girls Club of N.E. Texas", "Y.W.C.A. of Helena", "StepForward")
q  <- np_normalize(np_query(data.frame(unique_entity_id = seq_along(nm), name = nm,
        street = "", city = "", state = "OH", zip = "", stringsAsFactors = FALSE),
        c(.id="unique_entity_id", name="name", street="street", city="city",
          state="state", zip5="zip")))
data.frame(`Original name` = nm, `Cleaned match key` = q$name_key,
   `Tokens compared` = vapply(strsplit(q$name_key, " "), paste, "", collapse = " · "),
   check.names = FALSE)
```

Addresses are parsed the same way — a standardized street body, a separated
**unit**, and a **ZIP hierarchy** (the 5-digit ZIP nests inside a 3-digit region,
so a match can degrade gracefully when only partial geography is known).

```{r norm-addr}
ad <- data.frame(unique_entity_id = 1:3, name = c("A","B","C"),
   street = c("123 North Main Street, Suite 400","P.O. Box 1234","45 Elm St Apt 2B"),
   city = c("Saint Louis","Cincinnati","Bethesda"), state = "OH",
   zip = c("63101-2200","45201","20814"), stringsAsFactors = FALSE)
qa <- np_normalize(np_query(ad, c(.id="unique_entity_id", name="name", street="street",
        city="city", state="state", zip5="zip")))
data.frame(`Original address` = ad$street, `Street key` = qa$street_key,
   Unit = ifelse(is.na(qa$street_unit), "—", qa$street_unit),
   ZIP5 = qa$zip5, `ZIP region` = qa$zip3, check.names = FALSE)
```

**Corrupted records are *tolerated*, not silently fixed.** Rather than guess a
correction, the matcher compares the fuzzy *similarity* of the two strings, so a
typo or a spacing slip in the official record still scores as a near-match. A few
real BMF cases the fuzzy layer recovers:

```{r norm-fuzzy}
data.frame(
  `Source name` = c("CENTER FOR TRANSPORTATION AND THE ENVIRONMENT",
                    "STEP FORWARD", "HILLEL FOUNDATION"),
  `BMF record (as filed)` = c("...AND THE ENVIROMENT  (missing N)",
                    "STEPFORWARD  (de-spaced)", "HILEL FOUNDATION  (dropped L)"),
  `Name similarity` = c(0.99, 0.98, 0.91), check.names = FALSE)
```

## 3. The two-stage idea

npmatch is deliberately split into two stages:

> **Stage 1 (the algorithm)** surfaces and sorts the candidates — fast,
> consistent, cheap across millions of records.
> **Stage 2 (a human or an LLM)** confirms the ambiguous cases and gathers the
> extra evidence the algorithm can't see.

The three labels are the handoff. **YES** is confident enough to accept
automatically; **NO** is confident enough to reject; **MAYBE** means *"a match
probably exists, but I can't be sure on name and address alone — please look."*
Stage 1's job is not to be right about everything — it is to be **right when it
is confident and honest when it isn't**, so Stage 2's effort goes only where it
is needed.

## 4. The matching cascade (fastest checks first)

The multi-tier strategy runs cheapest-to-most-expensive and stops working on a
source org once it is confidently placed. Exact name-and-place agreement is a
fast hash lookup; the broad token search (comparing every shared distinctive
word) is far costlier, so it only runs on what is still unresolved. On the 1,502
benchmark cases, **the four exact passes settle ~88% of all YES matches before
the expensive token search runs at all.**

```{r funnel}
fn <- data.frame(
  stage = c("exact name","exact name↔DBA","token, same state","de-spaced token","token, cross-state"),
  yes   = c(845, 22, 109, 7, 1),
  cost  = c("cheap","cheap","costly","costly","costliest"))
fn$stage <- factor(fn$stage, levels = rev(fn$stage))
if (have("ggplot2")) {
  ggplot(fn, aes(stage, yes, fill = cost)) +
    geom_col(width = .7) +
    geom_text(aes(label = yes), hjust = -0.2, size = 3.4) +
    coord_flip() + expand_limits(y = 950) +
    scale_fill_manual(values = c(cheap="#2c7fb8", costly="#7fcdbb", costliest="#edf8b1")) +
    labs(title = "YES matches resolved at each cascade stage",
         subtitle = "Cheap exact passes run first and settle most matches",
         x = NULL, y = "matches auto-accepted", fill = NULL) +
    theme_minimal(base_size = 11) + theme(legend.position = "bottom")
}
```

## 5. Reading the output: candidate sets

For each source organization, npmatch surfaces the candidate EINs it considered
with everything needed to judge the fit — the candidate's name, a **Name
Similarity** (0–1), which **Name Version** matched, the combined **Full Match
Score**, and any veto that fired. Three real examples show the flavor and the
pathologies. *(The source organization is named above each candidate set so you
can see what it is being matched against.)*

**(a) A clear winner amid noise.** The blocking step casts a wide net, so weak
look-alikes appear — but the true match dominates and is auto-accepted (YES).

> **Source organization:** *The North Carolina Coalition Against Domestic Violence, Inc.*

```{r cand-a}
data.frame(
  Candidate = c("The North Carolina Coalition Against Domestic Violence",
                "North Carolina (DBA fragment)", "NC Guardian ad Litem Assn"),
  `Name Similarity` = c(0.95, 0.86, 0.67),
  `Name Version` = c("name", "dba", "token overlap"),
  `Full Match Score` = c(0.95, 0.59, 0.48),
  Why = c("near-exact name + state → the pick",
          "shares only the state words → noise",
          "a few shared words, different org → noise"), check.names = FALSE)
```

**(b) An organization vs. its foundation (a family of related EINs).** When both
the operating org and its affiliated foundation sit at the same place, the
foundation is *soft-vetoed* — surfaced, but held back from auto-accept so a
reviewer confirms which EIN is meant.

> **Source organization:** *New York Rural Water Association*

```{r cand-b}
data.frame(
  Candidate = c("New York Rural Water Association", "New York Rural Water Foundation"),
  `Name Similarity` = c(1.00, 1.00), `Name Version` = c("exact", "exact"),
  `Full Match Score` = c("0.96", "→ MAYBE"),
  Veto = c("—", "affiliate_suffix (Foundation)"),
  Why = c("the operating org → auto-accept",
          "the affiliate arm → routed to review, not auto-accepted"),
  check.names = FALSE)
```

**(c) A genuinely ambiguous cluster → MAYBE.** Several neighborhood organizations
share the distinctive words; none clearly *is* the source org, so the algorithm
refuses to guess and sends the set to review.

> **Source organization:** *Young Men's and Young Women's Hebrew Association of Washington Heights*

```{r cand-c}
data.frame(
  Candidate = c("Washington Heights Inwood Mask Blocc...", "Washington Heights & Inwood Devel...",
                "Chamber of Commerce of Washington Hts", "Washington Heights Inwood Preserv..."),
  `Name Similarity` = c(0.79, 0.85, 0.79, 0.75), `Name Version` = "token overlap",
  `Full Match Score` = c(0.78, 0.73, 0.69, 0.67),
  Why = c("top candidate, but not decisive","close second","related, weaker","related, weaker"),
  check.names = FALSE)
```

The recurring pathologies: **shared generic/geographic words** (a),
**parent/subsidiary/affiliate families** (b), and **dense clusters of
similarly-named local orgs** (c). Name and address alone can rank these but
cannot always decide them — which is the entire reason for MAYBE.

## 6. The three answers — how the cases flow

Here is where all 1,502 hand-verified benchmark cases land. **MAYBE is the real
decision point**: YES is mostly settled and NO is mostly correct, but MAYBE is
where a real match hides behind ambiguity and human/LLM judgment pays off.

```{mermaid}
%%| label: fig-flow
flowchart TD
    A["<b>All source orgs</b><br/>n = 1,502"]
    A --> Y["🟢 <b>YES</b> · 65.2%<br/>auto-accept<br/>99.6% correct · 0.4% false-pos."]
    A --> M["🟡 <b>MAYBE</b> · 15.8%<br/>→ Stage 2 review"]
    A --> N["🔴 <b>NO</b> · 18.9%<br/>85.9% correct reject · 14.1% false-neg."]
    M --> M1["Confirmed match<br/>61%"]
    M --> M2["No match<br/>31%"]
    M --> M3["Needs more info<br/>8%"]
    N --> N1["Not a nonprofit<br/>43%"]
    N --> N2["Not in active BMF<br/>34%"]
    N --> N3["False negative<br/>14%"]
    N --> N4["Undetermined<br/>9%"]
    classDef yes fill:#c7e9c0,stroke:#31a354,color:#111;
    classDef maybe fill:#fee391,stroke:#d95f0e,color:#111;
    classDef no fill:#fdcfcf,stroke:#d73027,color:#111;
    class Y,M1 yes;
    class M,M2,M3 maybe;
    class N,N3 no;
```

**First pass — the algorithm's initial tier.** YES and NO are *decisions*; MAYBE
is *deferred* to Stage 2 (its resolution is the second table).

```{r first-pass}
data.frame(
  Branch = c("🟢 YES","🟡 MAYBE","🔴 NO"),
  `% of all cases` = c("65.2%", "15.8%", "18.9%"),
  `% correct` = c("99.6% — right pick", "— deferred to review", "85.9% — right rejection"),
  `% error` = c("0.4% — false positive", "—", "14.1% — false negative"),
  check.names = FALSE)
```

**Second pass — what MAYBE resolves to after review**, and how the NO branch
breaks down by reason. MAYBE is where the payoff is: **61% of it is a real match**
that only a second look confirms.

```{r maybe-resolve}
data.frame(
  `MAYBE resolves to` = c("Confirmed match", "No match", "Needs more info"),
  `% of MAYBE` = c("61.3%", "31.1%", "7.6%"),
  Meaning = c("reviewer / LLM confirms a real EIN",
              "no real match — surfaced for safety, then cleared",
              "needs external evidence to decide"), check.names = FALSE)
```

```{r no-types}
data.frame(
  `NO breaks into` = c("Not a nonprofit (govt / for-profit / person)",
                       "Nonprofit, not in the active BMF (foreign / new / co-op / church)",
                       "False negative — a real match was missed",
                       "Undetermined"),
  `% of NO` = c("43.3%", "33.5%", "14.1%", "8.5%"),
  check.names = FALSE)
```

### 🟢 YES — "accept this match" · **~98% correct (~1–2% false positive)**
Strong name and address agreement, no veto. Reviewer effort is light: confirm the
obvious pick. *Why the ~2% go wrong:* two different organizations sharing
distinctive words **and** an address —

- **Same campus, different entity** — a booster club, auxiliary, or foundation
  at its parent's building ("Virtua Health" vs "Virtua Health and Rehabilitation").
- **Same distinctive name, wrong entity type** — "Luzerne County Housing
  Authority" pulled toward "Luzerne County Historical Society." *(A new
  government-entity rule now routes most of these to MAYBE.)*

### 🟡 MAYBE — "a match likely exists; please review" · **the review tier**
Of everything sent to MAYBE, **61% turn out to be a real match**, 31% resolve to
no-match, and 8% need more information. Combined with YES, **~98% of all true
matches are surfaced** for a reviewer to see. *Why a case lands here, not in YES:*

- **Subset / superset names** — "Nowlin Hall" vs "Nowlin Hall Apartments": same
  org or a related property? Name alone can't tell.
- **Parent vs subsidiary vs chapter** — all share the distinctive words, so the
  algorithm surfaces them but refuses to guess which EIN.
- **Cross-state parents** — a national org filing under a headquarters in a
  different state than the local address.

### 🔴 NO — "no acceptable match found" · **mostly correct; ~2% of true matches missed**
Most NO answers are right. Among the correct rejections, **~50%** are genuinely
not nonprofits (government, for-profit, individuals), **~39%** are real nonprofits
absent from the *active* file (foreign, brand-new, co-ops, churches), and ~10%
are undetermined. The small share of **false negatives** happen two ways:

1. **The right candidate was never surfaced (a "blocking" miss)** — the names are
   too different for the search to connect:
   - **Rebrands** — "Coalition for the Common Good" is Antioch University's new
     legal name; the strings share nothing.
   - **Acronyms** — "YWCA of Helena" vs "Young Women's Christian Association."
   - **Typos in the record** — "…Environment" vs the BMF's "…Enviroment."
   - **Spacing / abbreviation** — "Step Forward" vs "StepForward." *(Now recovered
     automatically.)*
2. **The org truly isn't in the file** — churches (auto-exempt, often unlisted),
   foreign orgs, brand-new nonprofits. Here NO is *correct*; the reviewer just
   confirms it.

## 7. What a reviewer does in each tier

```{r workflow}
data.frame(
  Tier = c("🟢 YES","🟡 MAYBE","🔴 NO"),
  Effort = c("Light / spot-check","Moderate","Targeted"),
  Task = c("Confirm the auto-selected EIN against the surfaced candidates.",
           "Pick the right candidate, or gather a little evidence to decide.",
           "Confirm the org is genuinely unmatched — or research a possible miss."),
  check.names = FALSE)
```

Because YES is ~98% precise and NO is mostly correct, **the real work
concentrates in MAYBE** — exactly where you want human judgment spent.

## 8. Humans or LLMs — interchangeable at Stage 2

Stage 2 is the same task whether a person or a language model does it: *look at a
surfaced organization, decide whether a candidate is the right EIN, and if not,
find out what the organization actually is.* npmatch is built so the two are
drop-in substitutes.

To make the LLM path turnkey, the project ships a **tiered research protocol with
a reusable per-organization prompt** (`dev/RESEARCH-PROTOCOL.md`) that escalates
only as far as needed and is cost-aware:

- **Tier 1** — a free local recheck against the full IRS file (name variants,
  de-spaced forms, typos).
- **Tier 2** — nonprofit registries (e.g. ProPublica) for the EIN.
- **Tier 3** — open web search to establish what the org is when registries are silent.

Each organization returns a determination, the EIN found (if any), the source, a
confidence level, and a token/cost estimate — the same structured verdict a human
enumerator records, so the two mix freely (LLM triage first, humans on the
residual). This is what lets the review stage scale from hundreds of cases to
hundreds of thousands.

## 9. Where things stand today

- **YES precision ≈ 98%** — auto-accepts you can trust.
- **Coverage ≈ 98%** — of orgs that have a real match, ~98% are surfaced as YES
  or MAYBE; only ~2% are missed into NO.
- **MAYBE is ~16% of cases** — the focused Stage-2 queue.
- Matching against the **unified** (active + inactive) IRS file recovers a further
  ~6% of matches to historical / defunct EINs, each tagged `active` so downstream
  can filter current-990 matches from identity-resolution ones.

## Appendix — a peek at the evaluation dataset

Each row is one candidate the algorithm surfaced, with the fields a reviewer uses
to decide. (See the companion *Candidate Evaluation Frame* vignette for full
definitions.)

```{r preview}
data.frame(
  source_org = c("Coalition for the Common Good","YWCA of Helena","Nowlin Hall Apartments",
                 "Luzerne County Housing Authority","Freedom Bird Foundation"),
  best_candidate = c("(Antioch University — rebrand, not surfaced)","(Young Women's Christian Assn — acronym)",
                     "Nowlin Hall","Luzerne County Historical Society","Operation Freedom Birds"),
  name_sim = c(NA, NA, 0.95, 0.90, 0.72),
  tier = c("NO","NO","MAYBE","YES→(now MAYBE)","MAYBE"),
  gt_ein = c("31-0536640","81-0235416","23-7110959","(no match)","87-2482496"),
  note = c("rebrand: names share nothing (blocking miss)","acronym vs spelled-out (blocking miss)",
           "subset name → review","govt entity, wrong type (veto now catches)","short-form name → review"),
  stringsAsFactors = FALSE)
```
