---
title: "npmatch: The Training & Evaluation Dataset"
format:
  html:
    df-print: kable
    highlight-style: tango
    toc: true
    toc-location: right
    css: clean.css
    embed-resources: true
execute:
  echo: false
  warning: false
vignette: >
  %\VignetteIndexEntry{npmatch: The Training & Evaluation Dataset}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

*A hand-labeled benchmark of 1,502 organizations linking the SAM/USASpending
nonprofit registry to the IRS Business Master File — how it was built, how the
matcher performs on it, and how to use it to train and benchmark Stage-1
classifiers or drive an LLM review loop.*

## 1. What this dataset is

`npmatch` links organizations that appear as nonprofits in the SAM /
USASpending registry (keyed by **UEI**) to the IRS Business Master File of
exempt organizations (keyed by **EIN**). To measure and improve that process we
hand-labeled a benchmark of **1,502 source organizations**:

- **1,000 a true random sample** of SAM nonprofits — use these for *population*
  performance and outcome-class prevalence.
- **502 hard cases** — an oversample of the difficult residual (fuzzy names,
  DBA/address divergence, ambiguous entities). Use these to probe failure modes
  and train a classifier on the cases that actually matter; **do not** read
  population rates off this stratum.

The benchmark does triple duty: a **validation set** for the matcher, a
**training set** for Stage-1 (name/address) classifier models, and a **demo
corpus** for exploring the package.

## 2. How the dataset was built

```{mermaid}
flowchart LR
  A["SAM monthly extract<br/>(~165k flagged nonprofits)"] --> B["sample:<br/>1,000 random + 502 hard"]
  B --> C["np_cascade()<br/>vs unified BMF (3.69M)"]
  C --> D["YES / MAYBE / NO<br/>+ candidate scores"]
  D --> E["ground truth:<br/>algo + human + LLM + web"]
  E --> F["labeled benchmark<br/>(1,502 orgs)"]
```

1. **Source.** SAM's public monthly extract, filtered to nonprofit
   business-type codes (`np_flag_nonprofits()`; codes A8/BZ/2U/A7) — roughly
   165,000 organizations.
2. **Sample.** A 1,000-org random draw plus a 502-org oversample of hard
   residual cases, each tagged with `sample_type`.
3. **Match.** The shipped cascade (`np_cascade()`) against the **unified** BMF
   (active + inactive, 3.69M records), producing a YES/MAYBE/NO tier and scored
   candidates for every org.
4. **Ground truth.** Each case was adjudicated with a mix of algorithmic
   agreement, human review, LLM judgment, and web verification — confirming that
   accepted matches are correct, and (crucially) investigating whether
   *non*-matches are truly not nonprofits or simply missing from the BMF.

Each org is assigned one **outcome class**:

```{r}
data.frame(
  `Outcome class` = c("in_bmf_match", "nonprofit_not_in_bmf", "not_a_nonprofit",
                      "no_match_or_na", "cant_determine"),
  Meaning = c(
    "A correct BMF EIN exists (active or inactive).",
    "A real nonprofit with no BMF record (many churches, lapsed orgs, state-only registrations).",
    "Not actually a nonprofit — a SAM mislabel (for-profit, government unit).",
    "No match and not clearly classifiable as the above.",
    "Insufficient evidence to decide."),
  check.names = FALSE)
```

### The inactive-BMF correction

An important labeling subtlety: matching uses the **unified** BMF (active +
inactive) on purpose. 30 organizations first tagged "not in BMF" — because
labeling initially checked only the *active* file — turned out to be exact
name + address matches to **inactive** BMF records (lapsed or recently-registered
orgs). These were corrected to `in_bmf_match` with the inactive EIN (20 by an
exact name+address rule, 10 confirmed on manual/web review). A further 6
plausible-but-unconfirmed candidates were reviewed and rejected (affiliate
siblings such as a hospital's *Foundation*, and cross-state name collisions).
The `gt_notes` column records every such decision.

## 3. Record-linkage performance

The positive event is **a correct EIN link**. *Recall* = correct links ÷
linkable orgs; *precision* = correct ÷ accepted; the MAYBE tier is a human/LLM
review queue, not an error bucket.

```{r}
data.frame(
  Metric = c("Linkable orgs", "Non-linkable orgs", "Recall — auto (YES only)",
             "Recall — auto + review", "Precision (YES picks correct)",
             "Auto-accept false positives"),
  `Random (n=1,000)` = c("807", "193", "88.1%", "95.0%", "99.2%", "5"),
  `Hard (n=502)`     = c("380", "122", "70.0%", "86.1%", "96.0%", "7"),
  `All (n=1,502)`    = c("1,187", "315", "82.3%", "92.2%", "98.3%", "12"),
  check.names = FALSE)
```

Read the **random** column as expected real-world performance: ~95% of linkable
nonprofits are recovered once the review tier is worked, at ~99% precision on the
auto-accepted matches. The hard column is the deliberately-difficult residual.

**Algorithm-only tiering** (before any review):

```{r}
data.frame(
  Tier = c("YES (auto-accept)", "MAYBE (review)", "NO (withhold)", "No candidates"),
  Random = c("717", "123", "144", "16"),
  Hard   = c("277", "136", "87", "2"),
  Meaning = c("strong name + address, no veto", "surfaced for human/LLM confirmation",
              "nothing cleared the review floor", "blocking found nothing"),
  check.names = FALSE)
```

Because some organizations are legitimately unmatchable, the match rate must be
read against how often a match *can* exist. From the representative random
sample:

```{r}
data.frame(
  `Outcome class` = c("in BMF (matchable)", "nonprofit, not in BMF",
                      "not a nonprofit (SAM mislabel)", "no match / n.a.", "can't determine"),
  Random = c("807", "85", "65", "22", "21"),
  Share  = c("80.7%", "8.5%", "6.5%", "2.2%", "2.1%"),
  check.names = FALSE)
```

So ~81% of SAM "nonprofit" registrations are matchable to the BMF; the ~19%
remainder are correctly *not* matched — and a NO here can be a **true negative**
(the record is not in fact a nonprofit), not a failed match. Two denominators
therefore matter: **recall among the linkable ~81%** (~95%), and **yield over the
whole population** (~77% of all source orgs receive a correct link).

## 4. The dataset files

```{r}
data.frame(
  File = c("EVAL-FRAME-1502-MERGED.csv", "GROUNDTRUTH-MASTER-1502.csv",
           "TRAINING-PAIRS-1502.csv", "DATA-DICTIONARY-EVAL-FRAME.md / .csv"),
  Grain = c("candidate pair", "source org", "candidate pair", "—"),
  Rows = c("2,763", "1,502", "304,274", "150"),
  Description = c("Reviewable evaluation frame (150 cols) + ground truth.",
                  "One row per org: the archival label key.",
                  "Model-training projection: features + label per scored pair.",
                  "Column dictionary for the evaluation frame."),
  check.names = FALSE)
```

The three data frames are three *views* of the same run: the **ground-truth
master** is the label key; the **evaluation frame** joins those labels onto the
reviewable top candidates; the **training pairs** are a projection of every
scored candidate (features + a boolean `label`).

## 5. Benchmarking Stage-1 classifiers

`TRAINING-PAIRS-1502.csv` is one row per scored candidate pair with a boolean
`label` (1 = this EIN is the true match) and the comparison features the scorers
calibrate on: the continuous field similarities **`name_sim`**, **`street_sim`**,
**`city_sim`**, **`zip_sim`** (Jaro-Winkler / agreement, 0–1), the binary
geographic-agreement flags **`geo_stnum`/`geo_zip9`/`geo_zip5`/`geo_zip3`/
`geo_state`/`geo_pobox`**, the categorical **`name_match_type`**, the name
distinctiveness **`name_freq`**, and the current blended **`score`**. It is
exactly the input to the package's model tools:

```{r}
#| echo: true
#| eval: false
library(npmatch)
path  <- np_fetch("https://<archive>/TRAINING-PAIRS-1502.csv")   # fetch-only (too large to bundle)
pairs <- read.csv(path)

# cross-validated comparison of weighted / EM (reclin2) / model scorers
np_benchmark(pairs, methods = c("weighted", "em", "model"),
             target_precision = 0.98, min_recall = 0.95)

# fit a single Stage-1 model on the field-similarity features
fit <- np_train(pairs, engine = "logit",
                features = c("name_sim", "street_sim", "city_sim", "zip_sim",
                             "geo_stnum", "geo_zip9", "geo_zip5", "geo_state", "geo_pobox"))

# score quality + a precision-first operating point
np_auc(pairs$score, pairs$label)
np_tune_thresholds(pairs$score, pairs$label,
                   target_precision = 0.98, min_recall = 0.95)
```

On this benchmark the geography-adaptive **hier** scorer is the shipped default
(it beat a cross-validated GLM at scale); the training pairs let you reproduce
that comparison or fit your own Stage-1 classifier.

## 6. What the evaluation dataframe looks like

`EVAL-FRAME-1502-MERGED.csv` is **candidate-level**: one row per (source org ×
BMF candidate). Group by `uei` to see an org's competing candidates. Two flags
orient every group:

- `is_top_candidate = 1` — the algorithm's chosen pick.
- `is_gt_ein = 1` — the row whose EIN is the ground-truth correct match.

Alongside the similarity features it carries the full evidence a reviewer needs:
normalized name variants (`name_*_normalized`), IDF-annotated tokens
(`name_*_tokenized`), geographic agreement flags (`geo_*`), the veto trail
(`veto*`), and imported BMF context (`BMF_ntee_clean`, `BMF_subsection`,
`BMF_form_990`, …) and SAM context (`SAM_bus_type_*`). All 150 columns are
documented in `DATA-DICTIONARY-EVAL-FRAME.md`. This is the frame the
[candidate-evaluation-frame](candidate-evaluation-frame.html) vignette describes
field by field.

## 7. From MAYBE to an LLM review loop

The MAYBE tier is where the algorithm defers to judgment. `np_route()` turns a
cascade result into a review packet, and `np_as_prompt()` / `np_as_prompts()`
render each MAYBE case as a self-contained LLM adjudication prompt — the source
org, its candidate BMF entities with fields and similarities, why it was flagged,
and a JSON-output instruction:

```{r}
#| echo: true
#| eval: false
routing <- np_route(res, bmf = bmf_raw, sam = sam_raw)   # res = np_cascade() result
prompts <- np_as_prompts(routing)                        # one prompt per review case

# deploy: send each prompt to your LLM, collect {ein, decision, confidence, reason}
for (id in names(prompts)) {
  verdict <- my_llm(prompts[[id]])                       # your model call
  # ...record verdict, promote confirmed matches, drop rejected ones
}
```

The same packet drives a human reviewer or an LLM — the prompt is just formatted
text, so the second-stage process is model-agnostic and easy to deploy. This is
the two-stage philosophy in practice: the algorithm surfaces and sorts; a
human or LLM confirms the uncertain middle.

## 8. Loading the data

The evaluation frame and the ground-truth master ship **with the package** and
lazy-load by name. The full training pairs (304k rows) are too large to bundle,
so they are fetched from the shared data archive on demand.

```{r}
#| echo: true
#| eval: false
data("npmatch_eval")          # candidate-level evaluation frame + ground truth (2,763 x 150)
data("npmatch_groundtruth")   # query-level labels (1,502 x 17)

# full training pairs: fetch from the shared archive
path  <- np_fetch("https://<archive>/TRAINING-PAIRS-1502.csv")
pairs <- read.csv(path)
```
