---
title: "npmatch: Workflow & Project Structure"
format:
  html:
    theme: readable
    df-print: kable
    highlight-style: tango
    toc: true
    toc-location: right
    css: clean.css
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{npmatch: Workflow & Project Structure}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(npmatch)
knitr::opts_chunk$set(comment = "#>")
```

*How a matching run is organized on disk, and the order of operations that turns
a raw source extract into a finished crosswalk. This is the operational
companion to the [Reviewer's Guide](reviewer-guide.html): that article explains
what the matcher decides, this one explains where everything lives and which
function to call next.*

## 1. The workflow at a glance

A run is **one directory**, scaffolded up front and filled in as you go. Six
numbered stage folders, each producing files with the same shape, ending in a
crosswalk.

### 1.1 One directory per run

```{r init}
run <- file.path(tempdir(), "2026MAY")
np_project_init(run, run_id = "2026MAY", quiet = TRUE, use = TRUE)
```

That single call creates every folder below. The files marked **`*`** are the
ones it *doesn't* create — those arrive later, as each stage runs. An
empty-looking project is the normal starting state:

```
runs/2026MAY/
├── README.md · AGENTS.md · config.yml · manifest.csv · RUN-STATUS.md
│
├── 00_bmf/        SOURCE.md            *   pointer to the reference vintage
├── 00_sams/       sam_query.csv        *   the source, filtered to nonprofits
│
├── 01_stage1/     stage1_yes.csv       *
│                  stage1_maybe.csv     *   ──▶ 02_stage2
│                  stage1_no.csv        *   ──▶ 03_stage3
│
├── 02_stage2/     stage2_yes.csv       *
│                  stage2_no.csv        *   ──▶ 03_stage3
│
├── 03_stage3/     stage3_yes.csv       *
│                  stage3_no.csv        *
│
└── 04_final/      crosswalk.csv        *   the deliverable
                   eval_frame.csv       *
```

Each stage folder also carries a `README.md`, a stats report, and three
subdirectories — `batches/`, `interim/`, `logs/` — expanded in the sections
below. All six folders are created unconditionally. **Deploying a stage is
optional**: one that never runs simply leaves its outputs absent, which
`np_project_status()` reports as *not run* rather than as an error. A
stage-1-only run is a legitimate run.

### 1.2 Order of operations

Everything flows down. Each stage takes one tier from the stage above it, and
every branch eventually lands in `04_final`.

```{mermaid}
%%| label: fig-waterfall
%%| echo: false
flowchart TD
    BMF["<b>00_bmf</b><br/>reference pointer"]
    SAM["<b>00_sams</b><br/>sam_query.csv"]
    S1["<b>01_stage1</b> · np_stage1_run()<br/>block → compare → score → veto → select → tier"]
    BMF --> S1
    SAM --> S1
    S1 --> Y1["🟢 <b>YES</b><br/>stage1_yes"]
    S1 --> M1["🟡 <b>MAYBE</b><br/>stage1_maybe"]
    S1 --> N1["🔴 <b>NO</b><br/>stage1_no"]
    M1 --> S2["<b>02_stage2</b> · np_stage2_run()<br/>LLM adjudication"]
    S2 --> Y2["🟢 <b>YES</b><br/>stage2_yes"]
    S2 --> N2["🔴 <b>NO</b><br/>stage2_no"]
    N1 --> S3["<b>03_stage3</b> · np_stage3_run()<br/>entity gate → LLM research"]
    N2 --> S3
    S3 --> Y3["🟢 <b>YES</b><br/>stage3_yes"]
    S3 --> N3["🔴 <b>NO</b><br/>stage3_no"]
    Y1 --> F["<b>04_final</b> · np_final_run()<br/>crosswalk.csv · eval_frame.csv"]
    Y2 --> F
    Y3 --> F
    N3 -.->|"unmatched — eval_frame only"| F
    classDef yes fill:#c7e9c0,stroke:#31a354,color:#111;
    classDef maybe fill:#fee391,stroke:#d95f0e,color:#111;
    classDef no fill:#fdcfcf,stroke:#d73027,color:#111;
    classDef stage fill:#eeeeee,stroke:#777,color:#111;
    class Y1,Y2,Y3 yes;
    class M1 maybe;
    class N1,N2,N3 no;
    class S1,S2,S3,F,BMF,SAM stage;
```

Three things the picture is meant to make obvious.

**The stages are a funnel of decreasing automation and increasing cost.** Stage
1 decides everything it can from string similarity and geography alone, at
essentially no marginal cost per record. Stage 2 spends an LLM call on the cases
that similarity genuinely cannot separate. Stage 3 spends an LLM call *and* a
web search on the cases the matcher rejected. You pay for judgment only where
arithmetic ran out.

**A NO is not the end.** Stage 1's rejections and stage 2's rejections pool
together and get researched. That is where recall the matcher missed comes back
— either because the right EIN was never surfaced by blocking, or because the
organization genuinely is not a nonprofit and should be recorded as such.

**`04_final` sees everything**, not just the matches. Unmatched ids still land
in `eval_frame.csv` carrying their final answer, which is what makes the run
measurable rather than merely finished.

### 1.3 The functions

Seven calls cover an entire run:

| call | what it does | writes |
|---|---|---|
| `np_project_init()` | scaffolds the run directory | the tree above |
| `np_prepare_sam()` | filters the raw extract to nonprofits | `00_sams/sam_query.csv` |
| `np_stage1_run()` | blocks, compares, scores, vetoes, selects, tiers | `stage1_yes` / `_maybe` / `_no`, `stage1_k_candidates` |
| `np_stage2_run()` | **called twice** — LLM adjudicates the MAYBEs | `stage2_yes`, `stage2_no` |
| `np_stage3_run()` | **called twice** — LLM researches the NOs | `stage3_yes`, `stage3_no`, `stage3_research_findings` |
| `np_final_run()` | binds the stages together | `crosswalk.csv`, `eval_frame.csv` |
| `np_project_status()` | scans the directory, reports what has run | `RUN-STATUS.md` |

They fit together through three conventions, and once you have these three the
rest of this article is detail.

**Every stage function defaults its paths into the active project.** `np_project_init()`
sets `options(npmatch.project=)`, so `np_stage2_run()` with no arguments already
knows where `stage1_maybe.csv` is and where to put its own output. You name paths
only when you deliberately want something outside the run.

**Stages 2 and 3 are called twice** because a reviewer sits in the middle. The
first call writes the work out for an agent and stops; the second collects what
came back. Between those two calls the run is idle on disk, in a state you can
inspect, hand to someone else, or resume tomorrow.

**Every stage writes the same columns.** `stage1_yes`, `stage2_no` and
`stage3_yes` are the same shape, so `np_final_run()` is a plain `rbind()` and any
stage can be skipped without the rollup noticing. See [§4](#sec-schema).

## 2. Building the data assets

Before any of that runs, the reference data has to exist. npmatch keeps data out
of the package, in **two separate places** that are easy to confuse:

| | shared asset store | run project |
|---|---|---|
| **managed by** | `np_data_*()` | `np_project_*()` |
| **holds** | reference vintages and expensive derived tables | one run's inputs, stage outputs, reports, logs |
| **how many** | one per machine | one per run |
| **default location** | `~/npmatch-data` | wherever you put it |

The BMF is roughly 3.5 GB raw plus a ~470 MB normalized bundle, and it is reused
by every run. Copying it per run is not an option, so the two are joined **by
pointers, not copies**. Build the store first, then point a run at it.

### 2.1 Create the store

```r
np_data_init()          # creates raw/ normalized/ results/ + MANIFEST.csv
np_data_root()          # where it landed
```

Three tiers, and the distinction is about how expensive each is to lose:

- **`raw/`** — immutable source downloads (BMF, SAM extracts), one file per
  vintage. Re-downloadable, but slowly.
- **`normalized/`** — derived assets built once per vintage: the normalized
  reference bundle, the token-IDF table, the name-frequency table. Cheap to
  rebuild in principle, expensive in wall-clock.
- **`results/`** — outputs that outlive a single run.

`MANIFEST.csv` at the root is what makes a run reproducible: one row per asset
recording where it came from, when it was fetched, its MD5 and its row count.
Record every input as you fetch it.

```r
np_fetch_bmf("unified", dest = np_data_path("raw", create = TRUE))
np_manifest_add("bmf-unified-2026-01", tier = "raw",
                source = "NCCS BMF catalog",
                path   = np_data_path("raw", "bmf_unified_2026-01.csv"))
np_manifest()
```

The store's location resolves from `getOption("npmatch.data")`, then the
`NPMATCH_DATA` environment variable, then the `~/npmatch-data` default.

### 2.2 What normalization actually does {#sec-normalize}

The expensive asset in `normalized/` is the reference put through
`np_normalize()`. This is where most of the matching leverage comes from: the
scorer only sees fields this step produced, so anything not fixed here has to be
recovered later by a human or an LLM.

**Cleaned onto a common form** — the goal is parity with the BMF's own
`org_name_join` field, so both sides are damaged identically:

- uppercased and de-accented (`Latin-ASCII`), so `Muñoz` and `Munoz` agree
- `&` expanded to ` AND `
- **apostrophes deleted, not spaced.** The BMF holds `ST LUKES HOSPITAL`, having
  stripped the apostrophe by deletion. Turning it into a space on the query side
  only gives `ST LUKE S HOSPITAL`, and no possessive name can then clear an
  exact-name pass. This single decision is worth real recall: 3.5% of 990-PF
  grantee names carry an apostrophe.
- **periods deleted between alphanumerics** (`F.I.N.D.` → `FIND`, `Y.W.C.A.` →
  `YWCA`) but kept as a word break after a word (`ST. LUKES` stays two tokens)
- every other punctuation mark becomes a token separator
- abbreviations standardized whole-word: `ASSOC`/`ASSN` → `ASSOCIATION`, `NATL` →
  `NATIONAL`, `INTL` → `INTERNATIONAL`, `CTR` → `CENTER`, `UNIV` → `UNIVERSITY`,
  `SVCS` → `SERVICES`, `ST` → `SAINT`, `MT` → `MOUNT`
- legal suffixes detected, recorded, then **stripped from the match key** — `INC`,
  `CORP`, `LLC`, `FOUNDATION`, `TRUST`, `SOCIETY`, `INSTITUTE` and the rest
  inflate similarity between unrelated organizations
- street types USPS-standardized (`STREET` → `ST`, `AVENUE` → `AVE`, `NORTH` → `N`)
- full state names mapped to two-letter codes; ZIP leading zeros restored

**Added as features** — the columns the scorer and the veto layer read:

- `name_key` (the match key), `dba_key`, `name_full` (suffix retained),
  `name_form` (the detected legal form: INC / LLC / CORP / …)
- `name_nums` — embedded digit tokens. Chapter, local, post and lodge numbers are
  *identifiers*, not noise: `UAW Local 32` and `UAW Local 45` are two entities.
- `name_ord` — ordinals folded to a canonical rank, so `FIRST` and `1ST` are the
  same marker and `First Baptist` ≠ `Second Baptist` is detectable
- `name_dir` — directional words canonicalized (`SOUTHWEST` → `SW`). Word forms
  only: a bare `N`/`S`/`E`/`W` is usually an initial or a street directional.
- `street_num`, `street_name`, `street_type`, `street_unit`, `is_po_box` — the
  address split into parts that can be compared independently
- `zip5`, `zip3`, `zip9` — a **nested prefix family** (`zip3` ⊂ `zip5` ⊂ `zip9`),
  so geographic agreement has granularity rather than being a coin flip
- `state_abb`

**Corpus statistics** built alongside, because distinctiveness is a property of
the reference, not of a pair:

- `np_name_freq()` — how many reference records share each exact `name_key`. An
  exact match to a name held by one organization is trustworthy without any
  address corroboration; an exact match to `FIRST BAPTIST CHURCH` is not.
- `np_token_idf()` — per-token `log(N / df)`. Rare tokens are what the name-token
  blocking leverages; corpus-common tokens are not.

One caveat when reading the data dictionary: `name_gen` / `name_gen_rank` (a
`JR`/`SR` person-name generation marker) are **vestigial**, inherited from
npmatch's person-matching lineage. Organizations do not have generations, nothing
consumes these columns, and they are retained only so existing labeled extracts
keep their layout. The organization-level equivalents are `name_nums`, `name_ord`
and `name_dir` above.

### 2.3 Index the assets into the run

The run does not take a copy of the reference — it records *which* vintage it
used and reads the bytes from the store. The SAM extract is genuinely per-run, so
that one does live inside the project:

```
runs/2026MAY/
├── 00_bmf/                       the reference — pointed at, never copied
│   ├── README.md
│   ├── SOURCE.md              *  vintage, resolved path, md5, row count
│   ├── SIGNATURE.txt          *  np_normalize_signature() hash of the cache
│   └── logs/
│
└── 00_sams/                      the source extract — genuinely per-run
    ├── README.md
    ├── raw/                   *  SAM_PUBLIC_MONTHLY.dat (you put this here)
    ├── sam_query.csv          *  np_prepare_sam(); what stage 1 reads
    ├── interim/
    └── logs/
```

```r
np_project_init("runs/2026MAY", run_id = "2026MAY",
                bmf = np_data_path("normalized", "bmf-unified-2026-01.rds"))

np_fetch_sam(dest = np_project_path("00_sams", "raw", create = TRUE))

np_prepare_sam(np_project_path("00_sams", "raw/SAM_PUBLIC_MONTHLY.dat"),
               out = np_project_path("00_sams", "sam_query.csv"))
#> SAM: 787,412 records -> 61,204 nonprofits (7.8%)
```

`np_prepare_sam()` reads the headerless pipe-delimited extract, names its 142
columns, keeps the nonprofit business-type codes, and writes `sam_query.csv` —
the file every later stage reads.

### Reference drift is a stop condition

A normalized bundle is only valid for the normalization code that produced it.
`np_normalize_signature()` hashes that code path, and the run stores the hash it
saw in `00_bmf/SIGNATURE.txt`. If the signature stops matching the cached bundle,
the cache has drifted from the current rules — everything in
[§2.2](#sec-normalize) may have changed underneath it. Rebuild before matching,
or the run is not reproducible. Stop; do not work around it.

## 3. Inside a run project

Every stage folder repeats the same three subdirectories, and the distinction
between them is the one operational rule worth memorizing:

```{r subdirs}
list.files(file.path(run, "01_stage1"), recursive = TRUE, include.dirs = TRUE)
```

- **`batches/`** — sharded work handed out and collected back. Entirely derived;
  safe to delete and rebuild.
- **`interim/`** — expensive but not deliverable: the cascade result and scored
  pairs (`res-NN.rds`, `pairs-NN.rds`). **Do not delete these until `04_final` is
  complete.** They are the only record of what blocking actually surfaced;
  without them the candidate sets cannot be rebuilt except by re-running the
  match against a possibly-changed reference.
- **`logs/`** — per-run logs. Write to the file, not only to the console.

`np_project_path()` takes a stage name — canonical (`"01_stage1"`), short alias
(`"stage1"`, `"bmf"`, `"final"`), or a bare number — plus an optional file below
it, so paths never have to be assembled by hand.

```{r paths}
old <- options(npmatch.project = "runs/2026MAY")

np_project_path("stage1", "stage1_yes.csv")
np_project_path("final",  "crosswalk.csv")
np_project_path("03_stage3", "batches/packets")

options(old)
```

Pointing a later session back at a run is `np_project_use("runs/2026MAY")`, which
sets the same option `np_project_init()` did.

## 4. One schema across every stage {#sec-schema}

Every `stage*_yes` / `stage*_maybe` / `stage*_no` file carries the same columns,
in the same order, with **exactly one row per source id**:

```{r schema}
np_stage_schema()
```

That uniformity is the point: the final rollup is a plain `rbind()`. The files
differ only in `outcome` (`YES` / `MAYBE` / `NO`) and in who decided —
`decided_by` is `algorithm` for stage 1, `llm_review` for stage 2 and
`llm_research` for stage 3, so a downstream consumer can always tell an
auto-accepted match from an adjudicated or researched one.

Anything finer-grained than one row per source id lives in a **side table joined
by `uei`**:

- `01_stage1/stage1_k_candidates.csv` — one row per surfaced candidate. This is
  where a NO's near-miss information lives: what it almost matched, and at what
  score. It is deliberately *not* in `stage1_no.csv`, which stays one row wide.
- `03_stage3/stage3_research_findings.csv` — the full research record, including
  the determination taxonomy (`match`, `not_a_nonprofit`, `nonprofit_not_in_bmf`,
  `cant_determine`) that `reason` carries in condensed form.

## 5. Stage 1 — the probabilistic cascade

```
01_stage1/
├── README.md
├── stage1_yes.csv             *  ──▶ 04_final
├── stage1_maybe.csv           *  ──▶ 02_stage2
├── stage1_no.csv              *  ──▶ 03_stage3
├── stage1_k_candidates.csv    *  one row per surfaced candidate
├── STAGE1-REPORT.md           *  per-pass cascade table, tier counts, timings
├── STAGE1-STATS.csv           *  per-chunk tally
├── batches/                   *  crosswalk-NN · review-NN · unmatched-NN · report-NN
│   └── shards/                *  review-sized parts handed to stage 2
├── interim/                   *  res-NN.rds · pairs-NN.rds — EXPENSIVE, keep
└── logs/                      *  run-<timestamp>.log · .err · progress.log
```

```r
np_stage1_run(reference = bmf, compute_size = 2500, review_size = 250, k = 3)
```

### What happens to a candidate pair

```{mermaid}
%%| label: fig-cascade
%%| echo: false
flowchart TD
    Q["<b>sam_query.csv</b><br/>one row per source organization"]
    B["<b>np_block()</b> — 7 passes, tight → loose<br/>exact name · exact DBA · shared rare token · cross-state<br/><i>queries resolved to YES drop out after each pass</i>"]
    C["<b>np_compare()</b> — reclin2 pairs and compares<br/>name_key similarity + the geo fields<br/>street · city · zip5 · zip3 · state"]
    S["<b>np_score()</b> — method = 'hier'<br/>0.60 × name + 0.40 × geo<br/><i>geo = max, not sum: zip9 ▸ street ▸ zip5 ▸ pobox ▸ zip3 ▸ city ▸ state</i>"]
    V["<b>np_veto()</b> — do-not-match rules"]
    X["pair removed from contention<br/><i>the query still gets a tier,<br/>from whatever candidate survives</i>"]
    SEL["<b>np_select()</b> — best candidate per query<br/>under three views: overall · name-only · addr-only"]
    T["<b>np_tier()</b>"]
    Q --> B --> C --> S --> V
    V -->|"<b>hard</b>: number · ordinal · direction · for-profit form"| X
    V --> SEL --> T
    T -->|"score ≥ 0.78, margin ≥ 0.05,<br/>no soft veto"| YES["🟢 <b>YES</b> — auto-accept"]
    T -->|"0.65 ≤ score < 0.78<br/>or near-tie with runner-up<br/>or <b>soft</b> veto: affiliate · government"| MAYBE["🟡 <b>MAYBE</b> — needs a reviewer"]
    T -->|"score < 0.65<br/>or nothing survived"| NO["🔴 <b>NO</b> — rejected"]
    classDef yes fill:#c7e9c0,stroke:#31a354,color:#111;
    classDef maybe fill:#fee391,stroke:#d95f0e,color:#111;
    classDef no fill:#fdcfcf,stroke:#d73027,color:#111;
    classDef step fill:#eeeeee,stroke:#777,color:#111;
    class YES yes;
    class MAYBE maybe;
    class NO,X no;
    class Q,B,C,S,V,SEL,T step;
```

**Blocking is a cascade, not one join.** Seven passes run tight to loose, and
after each pass the queries that reached YES are removed from the pending set, so
each successive, more expensive pass only processes the harder residual:

| # | pass | pairs on |
|---|---|---|
| 1 | `exact-name` | `name_key` == `name_key`, same state |
| 2 | `exact-name-dba` | `name_key` == `dba_key`, same state |
| 3 | `exact-dba-name` | `dba_key` == `name_key`, same state |
| 4 | `exact-dba-dba` | `dba_key` == `dba_key`, same state |
| 5 | `token-state` | a shared name token, same state (token IDF ≥ 8, token in ≤ 25,000 reference records) |
| 6 | `token-concat` | as above, also matching de-spaced compounds (`STEPFORWARD` ↔ `STEP FORWARD`) |
| 7 | `token-crossstate` | a shared name token, **any** state (token in ≤ 5,000 records) |

Candidates are scored once and the final `np_select()` / `np_tier()` run over the
de-duplicated union, so a match found by a loose pass still competes on equal
footing with one found early.

**The vetoes exist because string similarity is worst exactly where it matters.**
"First Presbyterian Church" and "Second Presbyterian Church" are ~95% similar and
are two different congregations. No amount of scorer tuning separates them: the
one differing token is drowned out by the agreeing ones. The rules encode that
token as a constraint on identity instead, and run *after* scoring so they can
override a strong match. Hard vetoes remove the pair; soft vetoes cap it at
MAYBE. `np_veto_audit()` shows what each rule is firing on.

**Two things demote a YES to MAYBE**: a near-tie with the runner-up (`margin <
0.05` — genuine ambiguity between two reference records), and a soft veto on the
selected pair. A soft veto never pushes a MAYBE down to NO.

### Two kinds of NO

A query for which blocking surfaced *nothing at all* is written to `stage1_no`
with `reason = no_candidates` and has no rows in `stage1_k_candidates`. That is a
different failure from a query whose candidates all scored below the floor, and
keeping the two separate is what makes blocking recall measurable — the first is
a blocking problem, the second a scoring problem, and they are fixed by different
work.

The run is **resumable**: a chunk counts as done when its per-chunk report
exists, and `only =` re-runs a specific subset. Read `STAGE1-REPORT.md` — the
per-pass table, tier counts and timings — before moving on.

## 6. Stage 2 — adjudicating the MAYBEs {#sec-stage2}

```
02_stage2/
├── README.md
├── PROMPT.md                     the agent's instructions (written by init)
├── stage2_yes.csv             *  ──▶ 04_final
├── stage2_no.csv              *  ──▶ 03_stage3
├── STAGE2-REPORT.md           *
├── batches/
│   ├── slim/                  *  slim-NN.csv       — questions out
│   └── decisions/             *  decision-NN.csv   — answers back
├── interim/
└── logs/
```

Two calls, with an adjudicator in the middle:

```r
np_stage2_run(shard_size = 250)   # 1st call: writes batches/slim/*.csv, stops
# ... point an agent at 02_stage2/PROMPT.md; it writes one
# ... decision-<shard>.csv per shard into 02_stage2/batches/decisions/
np_stage2_run()                   # 2nd call: collects, writes stage2_yes / stage2_no
```

### What a case looks like

Here is a real MAYBE from the shipped benchmark — the SAM registrant **Alston
Wilkes Society** of Columbia, SC, and the three candidates stage 1 surfaced for
it:

```{r stage2-case}
ex <- npmatch_eval[npmatch_eval$uei == "C5DVNAHN8F94", ]
rownames(ex) <- NULL

ex[, c("ein", "name_bmf_raw_main", "name_similarity", "addr_similarity",
       "total_score", "veto_soft_reason")]
```

Every candidate scores **0.98**. All three have identical name similarity,
identical address similarity, and sit at the same street address in Columbia.
The margin between first and second place is **0.00**:

```{r stage2-reason}
ex$decision_reason[1]
```

This is the case the arithmetic cannot decide, and the reason stage 2 exists. A
scorer sees three indistinguishable candidates; a reader sees a Society, its
Foundation, and its Association — a parent and two affiliates sharing a building.
Note that `affiliate_suffix` fired as a soft veto on the Foundation, which is what
capped the whole case at MAYBE instead of letting a coin-flip auto-accept.

The adjudicator gets the columns above plus the raw names on both sides, the
aligned address fields, and the BMF and SAM context columns, and answers one
question: *is one of these the same legal entity as the source?* It writes one
row per source id:

```
# 02_stage2/batches/decisions/decision-07.csv
uei,best_ein,llm_decision,llm_confidence,llm_reason
C5DVNAHN8F94,57-0477907,YES,high,"legal name matches exactly; Foundation and Association at the same address are its affiliates"
```

The ground-truth label for this case agrees — `57-0477907`, the Society itself.

The contract is fixed and `np_merge_decisions()` enforces it: `uei`, `best_ein`,
`llm_decision` (`YES`/`NO` only), `llm_confidence` (`high`/`medium`/`low`),
`llm_reason`. Every uei in the shard gets a row including the rejections, every
`YES` carries a `best_ein` drawn from that uei's own candidates, and every `NO`
leaves it empty.

### Partial returns

Re-running after only some shards come back collects what is there and reports
which are still outstanding. **An unreturned shard is a hole in the results, not
a set of NOs**, and is never silently treated as one. An empty MAYBE queue is a
legitimate outcome — stage 1 resolved everything — and writes empty outputs so
the stage reads as complete.

## 7. Stage 3 — researching the NOs

```
03_stage3/
├── README.md
├── PROMPT.md                        the agent's instructions
├── stage3_yes.csv                *  ──▶ 04_final
├── stage3_no.csv                 *
├── stage3_research_findings.csv  *  the full research record
├── STAGE3-REPORT.md              *
├── batches/
│   ├── packets/                  *  run-NNNN.md    — questions out
│   └── out/                      *  run-NNNN.tsv   — answers back
├── interim/                      *  stage3_seed.csv, cache/
└── logs/
```

Same two-call shape. By default it pools the NO cases from both stage 1 and
stage 2; pass `sources = "stage1_no"` to research only what the matcher rejected
outright.

```r
np_stage3_run(source_data = sam_query, bmf_index = bmf[, c("ein", "name")])
# ... agent follows 03_stage3/PROMPT.md, writing run-<NNNN>.tsv into batches/out/
np_stage3_run()
```

**The gate settles what it can for free.** Before anything is looked up,
`np_entity_gate()` screens the pool: foreign registrants, for-profit legal forms,
government units and sole proprietors get a determination and high confidence
with no lookup at all. An escape hatch still researches a screened case where
there is contrary evidence — a hit in `bmf_index`, an unreliable SAM flag, or a
990 filer sharing the registrant's website.

**A match must resolve to an EIN this reference actually contains.** Passing
`bmf_index` enforces that on collection: an EIN that is real but absent from this
vintage is reclassified to `nonprofit_not_in_bmf` with the EIN retained, because
a match the crosswalk cannot join is not a match.

## 8. `04_final` — the rollup

```
04_final/
├── README.md
├── crosswalk.csv     *  every matched source id, one row each
├── eval_frame.csv    *  every candidate, annotated with the final answer
├── FINAL-REPORT.md   *  the end-to-end funnel
├── interim/
└── logs/
```

```r
np_final_run()
```

Stage 3 recovers matches the matcher never surfaced, so a final EIN is often
absent from `stage1_k_candidates`. Those rows are not dropped — a synthetic row
carries the EIN with its similarity columns empty, and `final_ein_in_candset`
records which case it is. That flag splits recall loss into **blocking failure**
(never surfaced; no threshold change recovers it) and **scoring failure**
(surfaced but under the floor; recoverable by recalibration). Collapsing the two
would hide the only number that says which to work on.

## 9. Knowing where you are

```{r status}
s <- np_project_status(run, write = TRUE, quiet = TRUE)
attr(s, "stages")[, c("stage", "title", "status", "present", "expected")]
```

`np_project_status()` is a **scanner, not a ledger**. It reads the directory
every time it is called, so it cannot drift out of sync with reality the way a
stored state file would: delete an output and the stage reports as incomplete on
the next call. `RUN-STATUS.md` at the project root is the written form of the
same scan, regenerated on every call.

A stage is `complete` when every expected output is present, `partial` when some
are, and `not run` when none are.

```{r status-detail}
head(s[, c("stage", "file", "exists")], 8)
```

```{r cleanup, include=FALSE}
unlink(run, recursive = TRUE)
options(npmatch.project = NULL)
```

## 10. Rules of the road

These are the conventions the layout assumes. `AGENTS.md` in every scaffolded
project repeats them for an agent driving the run.

- **Never edit a stage's outputs by hand.** Re-run the stage function. A
  hand-patched CSV cannot be reproduced, and it silently invalidates every count
  downstream of it.
- **Never delete `interim/`** until `04_final` is complete.
- Everything under `batches/` is derived — safe to delete and rebuild.
- Write logs to the stage's `logs/`, not to the console only.
- Record every input and output in `manifest.csv` via `np_manifest_add()`.
- **Stop rather than improvise** when: the signature does not match the cached
  reference; a stage's outputs exist but its row counts do not reconcile with the
  previous stage (`stage2_yes` + `stage2_no` should equal `stage1_maybe`); or a
  queued batch returns no output file.

## 11. The whole run

```r
library(npmatch)

# once per machine
np_data_init()

# once per run
np_project_init("runs/2026MAY", run_id = "2026MAY",
                bmf = np_data_path("normalized", "bmf-unified-2026-01.rds"))

np_prepare_sam("runs/2026MAY/00_sams/raw/SAM_PUBLIC_MONTHLY.dat",
               out = np_project_path("00_sams", "sam_query.csv"))

np_stage1_run(reference = bmf)
np_project_status()

np_stage2_run(); np_stage2_run()    # agent works between the two calls
np_stage3_run(); np_stage3_run()

np_final_run()
np_project_status()
```

## 12. The next vintage

A new source extract does not mean a new run from scratch. `np_diff_unmatched()`
takes a fresh extract and an existing crosswalk and returns the ids that are
genuinely new or still unmatched — the only ones worth paying to match again.

```r
fresh <- np_prepare_sam("path/to/SAM_PUBLIC_MONTHLY_2026_AUG.dat")
todo  <- np_diff_unmatched(fresh, crosswalk = "runs/2026MAY/04_final/crosswalk.csv")

np_project_init("runs/2026AUG", run_id = "2026AUG")
data.table::fwrite(todo, np_project_path("00_sams", "sam_query.csv"))
```

Each vintage gets its own project directory, and runs are never edited in place.
That is what makes it possible to say which reference vintage produced any given
row of any given crosswalk.

## See also

- [A Reviewer's Guide](reviewer-guide.html) — what the matcher decides, and where a human comes in.
- [The Training & Evaluation Dataset](training-dataset.html) — the hand-labeled benchmark.
- [Building a Classifier](building-a-classifier.html) — training and calibrating stage-1 scoring.
- [The Candidate Evaluation Frame](candidate-evaluation-frame.html) — the candidate-level frame `04_final` produces.
